\(Q\)-Boson model and relations with integrable hierarchies

### 2024-04-27 draft.tex
finish introduction and add definitions

\usepackage{/home/thiago/.config/dot-files/latex/globaldef}
\input{/home/thiago/.config/dot-files/latex/definitions.tex}

@@

\address{\noindent 
Instituto de Física Teórica, UNESP-Universidade Estadual Paulista,
R. Dr. Bento T. Ferraz 271, Bl. II, Sao Paulo 01140-070, SP, Brazil\\
\&
Instituto de Física, Universidade de S\~ao Paulo,
Rua do Matão Travessa 1371, 05508-090 São Paulo, SP. Brazil
}

@@

%\date{\today}
\keywords{Integrability, Bethe, Schur}
\subjclass[2020]{82B20, 82B23}
\maketitle

@@

@@

>>>>

\usepackage{globaldef}
\input{definitions.tex}

@@

\address{\noindent 
Instituto de Física Teórica, UNESP-Universidade Estadual Paulista,
R. Dr. Bento T. Ferraz 271, Bl. II, Sao Paulo 01140-070, SP, Brazil\\
\&\\
Instituto de Física, Universidade de S\~ao Paulo,
Rua do Matão Travessa 1371, 05508-090 São Paulo, SP. Brazil
}

@@

%\date{\today}
%\keywords{Integrability, Bethe, Schur}
\subjclass[2020]{82B20, 82B23}
\maketitle

@@

The exploration of connections between quantum and classical
exactly solvable models constitutes a significant endeavor aimed at
elucidating the overarching structure of integrable systems.  This
research avenue has yielded fruitful insights, as evidenced
by~\cite{Its:1992bj, Foda:2009zz, Alexandrov:2011aa, Araujo:2021ghu},
to name a few.  The present work is situated within this research field.

Here, we examine the emergence of classical integrable structures
within correlation functions of the q-boson system. This quantum
integrable system characterizes strongly correlated bosons confined to
a one-dimensional chain~\cite{Bogoliubov:1992, Bogoliubov:1997soj,
  Bogoliubov2005}. Notably, this model bears close relation to the AL
(Ablowitz-Ladik) model, which is an integrable discretization of the
nonlinear Schrödinger equation.

In~\cite{Bogoliubov2005}, the author demonstrates that the \(q = 0\)
limit of the q-boson, known as the \emph{phase model}, is associated
with the enumeration of plane partitions. This connection allows for a
Schur polynomials expansion for the Bethe states. Subsequently, this
result was extended to \(q > 0\) in~\cite{Tsilevich:2006}, where the
Bethe states are expressed in terms of Hall-Littlewood polynomials.

Foda and Wheeler~\cite{Foda:2008hn, Wheeler:2010vmq} utilize a rich
set of dualities between the ring of symmetric functions and
\(q\)-deformations of the Heisenberg algebra~\cite{Jing1991, Jing1995}
to explore the combinatorial properties of this spin chain. As a
consequence, it has been shown that the scalar product of off-shell
Bethe states are restricted tau functions of the KP
(Kadomtsev-Petviashvili) hierarchy.

This underlying combinatorial structure underscores the significance
of this model in various physical phenomena, notably in the context of
crystal melting and string theory~\cite{Okounkov:2003sp,
  Saulina:2004da}. The application of the q-boson model in these
contexts has been demonstrated in~\cite{Sulkowski:2008mx}, with the
string data recoverable as the chain length approaches infinity.

This paper builds upon these developments and aims to extend some of
these results. Firstly, we delve deeper into the connection between
the scalar product and tau functions, exploring them in a wider
context, particularly in the case of the Toda
hierarchy~\cite{Takasaki:2018wsv}. It is important to note that both
the AL and the KP hierarchies are reductions of the Toda
hierarchy. This paper takes steps towards resolving this problem,
which is instrumental in understanding the finer details of these
relationships.

Additionally, we explore other correlation functions within this model
and examine how they fit into the classical integrable context. It is
noteworthy that even when the object is not a tau function itself, it
may have an interesting expansion in terms of different classes of
Schur polynomials, highlighting some of its combinatorial
properties. We also investigate some of these expansions in this
paper.

In Section 2, we provide a review of the phase model and
q-bosons. This section serves to establish notation and emphasize the
key aspects relevant to our analysis.  Section 3 delves into the
analysis of the Phase Model and its connections with the KP and Toda
integrable hierarchies. We explore the general expansion of a phase
model correlation function that yields a Toda system solution.  In
Section 4, we shift our focus to the q-bosons. We discuss a
determinantal formula for the scalar product of two off-shell Bethe
states and its expansion in terms of big, supersymmetric Schur
functions, as well as in terms of Kostka-Foulkes polynomials.
Finally, in Section 5, we conclude with an overview of open problems.

@@

% Ablowitz-Ladik and Toda hierarchy - reduction -- understands this problem
